Tests of measurement invariance via stochastic processes: {S}ome properties

### 2013-08-07 frontiers/frontiers.Rnw
write simall-ordinal.R file. update frontiers.Rnw simulation section R code
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@231 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ (group A and group B) can be obtained by Maximum Likelihood(ML) from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,respectively. The LR test statistic for the given threshold $v$ is then 
\begin{equation}
\label{equ:LRT}
LR(v)=-2[\ell(\hat{\theta};x_{1},,,,x_{n})-{\ell(\hat{\theta}^
{(A)};x_{1},...,x_{m})+\ell(\hat{\theta}^{(B)};x_{m+1},...,x_{n})
}],
\end{equation}

@@

<<theta estimate>>=
# estimate theta 
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by calling the measurement invariance function in lavaan. 
<<LRT>>=
# LRT
LRT<-as.matrix(anova(theta, thetaAB))
print(LRT,na.print="", quote=FALSE)
@
\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ is a relatively-small number of categories and the sample size is not too large (LRT is sensitive to large sample and will return "false" significance results, details see next section). However, when $V$ (in this case, age) continuous variable, however, multiple-group analysis usually cannot be performed because there are no existing groups. Instead, we can follow the procedure proposed by Ed Merkle and Zeileis. Specifically, we can fit a model whose parameters are restricted to be equal across across all individuals and then examine how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$. To formalize the idea, we assume that the observations are ordered w.r.t. $V$ and define the $k$ dimensional cumulative score process as 
\begin{equation}
\label{equ:B}
B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ is some consistent estimate of the covariance matrix of the scores,e.g. information matrix. ~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t. $V$ and decorrelates the scores between $k$ components of the parameters in the model. 
Further, there exists a functional central limit theorem that allows us to make formal inference with this cumulative score process. Assuming that individuals are independent and the usual ML regularity condition hold, it is possible to show that(addref) 
\begin{equation}
B(\cdot;\hat{\theta})\overset{d}{\rightarrow} B^{0}(\cdot)
\end{equation} 
where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ is a $k$ dimensional Brownian bridge. Thus, we can construct tests of measurement invariance by comparing the behavior of cumulative score process to that of a Brownian bridge. In practice, there are three test statistics can be used: double maximum, CvM and maxLM. 
\begin{equation}
DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
\end{equation}
\begin{equation}
CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
\end{equation}
\begin{equation}
maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
\end{equation}

Specifically, we will use the previous example, except changing categorical agegroup to continuous age.
Like before, after read in data, we fit the model in lavaan package. 
<<theta estimate>>=
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we will use the gefp function in strucchange package to get the $B(\hat{\theta}_{ij})$. Then we can get the three test statistics' significance by using sctest function. 
<<Get Estimate>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval=rep(NA,3)
pval[1]<-sctest(gefp1,functional=maxBB)$p.value
pval[2]<-sctest(gefp1,functional=meanL2BB)$p.value
pval[3]<-sctest(gefp1,functional=supLM(0.1))$p.value
pval
@
In applying the proposed continuous test, the $p$ value for CvM and maxBB $p$ value is larger than likelihood test. Besides, CvM $p$ value is not significant.

@@

<<Get Estimate Alternative Way>>=
DM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="DM", fplot=TRUE) 
DM
CvM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="CvM", fplot=TRUE) 
CvM
#maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
#maxLM
@

@@

In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. Both $p-$values are clearly larger than that of the likelihood ratio test and neigher is significant at $\alpha=0.5$

\section{Simulation}

>>>>

The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ 
(group A and group B) can be obtained by Maximum Likelihood(ML) 
from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,
respectively. The LR test statistic for the given threshold $v$ is then

\begin{equation}
\label{equ:LRT}
LR(v)=-2[\ell(\hat{\theta};x_{1},,,,x_{n})-{\ell(\hat{\theta}^
{(A)};x_{1},...,x_{m})+\ell(\hat{\theta}^{(B)};x_{m+1},...,x_{n})
}],
\end{equation}

@@

<<theta estimate>>=
# estimate theta 
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by conducting 
log likelihood test, which is done by anova function in R. 
<<LRT>>=
# LRT
LRT<-as.matrix(anova(theta, thetaAB))
print(LRT,na.print="", quote=FALSE)
@
\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ 
is a relatively-small number of categories and the sample size 
is not too large (LRT is sensitive to large sample and will 
return "false" significance results, details see next section). 
However, when $V$ (in this case, age) continuous variable, however, 
multiple-group analysis usually cannot be performed because there are 
no existing groups. Instead, we can follow the procedure proposed by 
Ed Merkle and Zeileis. Specifically, we can fit a model whose parameters 
are restricted to be equal across across all individuals and then examine 
how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$. 
To formalize the idea, we assume that the observations are ordered w.r.t. $V$ and 
define the $k$ dimensional cumulative score process as 
\begin{equation}
\label{equ:B}
B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}
{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ 
is some consistent estimate of the covariance matrix of the scores,e.g. information matrix. 
~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t. 
$V$ and decorrelates the scores between $k$ components of the parameters in the model. 
Further, there exists a functional central limit theorem 
that allows us to make formal inference with this cumulative 
score process. Assuming that individuals are independent and the usual 
ML regularity condition hold, it is possible to show that(addref) 
\begin{equation}
B(\cdot;\hat{\theta})\overset{d}{\rightarrow} B^{0}(\cdot)
\end{equation} 
where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ 
is a $k$ dimensional Brownian bridge. Thus, we can construct tests of measurement 
invariance by comparing the behavior of cumulative score process to that of 
a Brownian bridge. In practice, there are three test statistics can be used: 
double maximum, CvM and maxLM. 
\begin{equation}
DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
\end{equation}
\begin{equation}
CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
\end{equation}
\begin{equation}
maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
\end{equation}

Specifically, we will use the previous example, except assuming ordinal agegroup to be continuous.
Like before, after read in data, we fit the model in lavaan package. 
<<theta estimate>>=
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we will use the gefp function in strucchange package to 
get the $B(\hat{\theta}_{ij})$. After that, we can get the 
three test statistics' significance by using sctest function.

<<Get Estimate>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval=rep(NA,3)
pval[1]<-sctest(gefp1,functional=maxBB)$p.value
pval[2]<-sctest(gefp1,functional=meanL2BB)$p.value
pval[3]<-sctest(gefp1,functional=supLM(0.1))$p.value
pval
@
In applying the proposed continuous test, the $p$ value for 
CvM and maxBB $p$ value is larger than likelihood test. 
Besides, CvM $p$ value is not significant.

@@

<<Get Estimate Alternative Way>>=
DM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="DM", fplot=TRUE) 
DM
CvM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="CvM", fplot=TRUE) 
CvM
#maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
#maxLM (Problem: nonconsistent with sctest(gefp1))
@

@@

In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. 
Both $p-$values are clearly larger than that of the 
likelihood ratio test and neither is significant at $\alpha=0.5$

\section{Simulation}
\section{Simulation 1}

In Simulation 1, we examined the extent to which the proposed tests
are sensitive to the specific model parameter that violates measurement
invariance.  If, say, a factor loading violates measurement
invariance, it is plausible that this violation impacts other
parameters such as factor covariances or other loadings 
associated with the factor in question.  Thus, the goal of the
simulation was to examine the extent to which the proposed tests
attribute the measurement invariance violation to the parameters that
are truly in violation.

\subsection{Methods}
To examine these specificity issues, we generated data from a
two-factor model with three indicators each (the same model used in
Merkle and Zeileis, \citeyearNP{MerZei13}). The measurement invariance
violation occurred in one of three places: the factor loadings
 $\lambda_{11}--\lambda_{62}$, the unique variance associated
 $\epsilon_{11}--\epsilon_{66}$, or the factor covariance $\sigma_{12}$.  
We then tested for measurement invariance in seven subsets of parameters: the three
individual parameters noted above, all six factor loadings $\lambda_{11}--\lambda_{62}$, 
all six unique variances $\epsilon_{11}--\epsilon_{66}$. The path diagram is shown in  
Figure~\ref{fig:path}.

\begin{figure}
\caption{Path diagram representing the factor analysis model 
         used for simulations.}
\label{fig:path}
\includegraphics{PathDiagram}
\end{figure}

Power and Type I error were examined across three sample sizes
($n=120,480,960$) , three numbers of categories ($m=4,8,12$)and
7 magnitudes of invariance violations. 
The meaurement invariance violations began at level $1+m/2$ of $V$
and were consistant thereafter. These
violations were based on the auxiliary variable $V$: individuals below
the $1+m/2$ percentile of $V$ deviated from individuals above the $1+m/2$ 
percentile  by $d$ times the parameters' asymptotic standard errors
(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,1.5$. 
For each combination of sample size ($n$) $\times$ violation
magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
generated and tested.Four statistics were examined above. In all conditions, we
maintained equal sample sizes at each level of the ordinal variable.

\subsection{Results}
<<sim1>>=
## seed for replication
RNGkind(kind = "default", normal.kind = "default")
set.seed(1163)

if(file.exists("sim1.rda")) {
  load("sim1.rda")
} else {
  sim1 <- simulation()
  sim1$nlevels <- factor(sim1$nlevels)
  levels(sim1$nlevels) <- paste("m=",levels(sim1$nlevels),sep="")
  sim1$nobs <- factor(sim1$nobs)
  levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
  levels(sim1$test) <- c("AIC","LM_uo", "LRT", "LRT_sb", "LM_o", "WDM_o",
                         "YB97")
  save(sim1, file="sim1.rda")
}
sim1$test <- factor(as.character(sim1$test),
  levels = c("LM_o", "WDM_o", "LRT", "LM_uo", "AIC", "LRT_sb", "YB97"),
  labels = c("maxLM_o", "WDM_o", "LRT", "LM_uo", "AIC", "LRT_sb", "YB97"))
@

\begin{figure}
\caption{Simulated power curves for the ordered and unordered
      $\max\mathit{LM}$ tests, the ordered double-max test, and the
      likelihood ratio test across three sample sizes $n$, three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
\label{fig:sim1res}
\setkeys{Gin}{width=\textwidth}
<<sim1res, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT","LM_uo"),]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@
\end{figure}
